A New Metropolis-Hastings Test for Fast Minibatch MCMC Inference and Optimization

### 2016-05-15 nips_2016.tex
Update to paper draft. More organized, new title, sections 1 and 2 in draft stage, more references.

\title{Fast Mini-Batch MCMC for Big Data Parameter Optimization}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods are one of the most popular techniques in machine learning
for Bayesian inference. With large datasets, however, MCMC methods can be intractably slow due to
the need to compute the likelihood of all the data items. Borrowing from the optimization
literature, one intuitive way to make MCMC faster is to compute necessary gradients or likelihood
ratios based only on a random subset of the data each iteration. Such stochastic MCMC has received
increased research attention in recent years, but it is relatively difficult to develop formal
guarantees for those algorithms. Some stochastic MCMC methods also involve adaptively increasing the
minibatch size, which can nullify the benefits of the method if too much data is used. In addition,
with large datasets, it is unclear if obtaining a full posterior distribution is necessary or better
than deriving just a single MAP or point estimate of the parameters. In this paper, we present an
alternative perspective on stochastic MCMC based on deriving high-quality point estimates of
parameters in the presence of big data. To this end, we present a different Metropolis-Hastings test
designed to be more efficient in the state space. We also augment the target posterior distribution
with a temperature parameter, alleviating the need to adjust minibatch sizes.  We theoretically
show the benefits of our approach, and verify with empirical benchmarks on synthetic data and neural
network optimization.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is one of the most popular techniques for Bayesian
computation. In general, an MCMC algorithm will propose samples from a proposal distribution $q$,
and decide whether to accept or reject the new sample based on a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. With traditional Bayesian posterior inference, the posterior
probability $p(\theta \mid x_1, \ldots x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ based
on i.i.d. data $\{x_i\}_{i=1}^n$ is the target distribution which one aims to approximate by
sampling. This formula is problematic when $n$ is large (e.g., on the order of billions) because the
likelihood must be evaluated based on each individual point.

We can use stochasticity~\cite{RobbinsMonro1951}.

In this paper, we resolve these challenges by proposing an MCMC algorithm for Bayesian parameter
estimation in which proposals are fast and mini-batch based. Our M-H test acceptance probability is
about 50 percent, reasonably high, which means our samples mix well.We experimentally show that our
approach is better than the state of the art.

Our goal is using MH testing is in the context of MCMC model optimization, especially for neural
networks. We are also interested in using it for other big data problems. We are not interested in
exact inference of the posterior model parameters for a very large dataset.

The example you have been using, the mixture of two gaussians, illustrates this point well. The
"full data" posterior  based on 300 points has two sharp modes. The "cutting the MH budget" paper
takes on average about half the data points to make steps on this distribution. That's intuitive.
But it is also not very useful. The more data you have, the sharper the modes based on all the data.
On a non-toy problem (e.g. > 10,000 points), the posterior would collapse to two points from which
the MCMC simulation could likely never escape. And the "cutting the budget" method would still
require almost all the data to make steps.

For any real big-data problem (billions of points) the story would be the same.

We should argue instead that for model parameter inference, the goal is either MAP or ML estimation.
In both cases the goal is to find a good posterior mode. But the exact form of the distribution is
arbitrary, and e.g. non-unit powers of the distribution are fine. That corresponds to changing the
temperature of the distribution, or changing the scale factor in log delta space (which the factor
N/n is doing).

So rather than fixing the posterior temperature and trying to use adjustable mini-batch size to
achieve it (which ends up reducing to nearly-full-batch updates) we are proposing to run at the
natural temperature induced by a given minibatch size. We can always increase the mini-batch size to
reduce the temperature. But there is no need to do this beyond the point where the posterior mode is
sharp and well-defined. That will occur at some fixed number of points (e.g. a few thousand,
depending on the problem), and far less than the full dataset.

\section{Related Work}\label{sec:related_work}

In this section, we review different results on improving MCMC methods with a focus on those dealing
with mini-batches of large datasets.

Hamiltonian Monte Carlo (HMC) methods attempt to simulate the ``physics'' of the
system\footnote{Intuitively, we can think of a probability distribution as being a distribution over
particles in the system, where particles are the values of random variables.} in order to generate
distant, high-quality sample proposals (those are the best kind of samples). Since they simulate the
physical dynamics of the system, total energy is preserved and their proposals are always
accepted and there is no need for a Metropolis-Hastings correction/test\footnote{I'm not sure why
this follows but I haven't read the details.}. Unfortunately, HMC methods require the use of a
gradient computation based on all of the data per iteration. (This means the advantage of HMC
methods is that they generate high-quality proposals that mix well; their downside is that they are
unsuitable for big data, which might involve Bayesian posterior inference over that data.)

The Stochastic Gradient Hamiltonian Monte Carlo (SGHMC) method~ \cite{sghmc_2014} tries to
compromise by using a \emph{mini-batch version} of HMC, so each iteration, only the mini-batch is
used to compute the gradient.  SGHMC also includes an extra friction term to counteract the effect
of noise on the system (or gradient) that is added due to using only a small subset of data. (The
``naive'' way of replacing the batch with the mini-batch \emph{without} friction means the system
deviates from the target/desired distribution.) ~\cite{sghmc_2014} assumes the injected noise is
Gaussian (we make a similar assumption) and appeals to the Central Limit Theorem and second-order
Langevin dynamics\footnote{There is a class of MCMC method techniques called \emph{Langevin
dynamics}~\cite{mcmc_hamiltonian_2010}. Those algorithms involve computing a gradient update for the
parameter of interest $\theta$ and then using the updated $\theta$ as the next sample (i.e., this is
like the proposal test). Moreover, by simulating the Hamiltonian dynamics of the system, there is no
need for a rejection test. To converge to a \emph{full (posterior) distribution} of $\theta$ and not
a single point (i.e., the MAP estimate of $\theta$, or the mode of the distribution) langevin
dynamics adds Gaussian noise each iteration to cause the $\theta$ samples to ``jump around.'' The
work of~\cite{langevin_2011} is closer to ours because it applies the normal Langevin dynamics
algorithm with one difference: the gradient updates are based on a \emph{mini-batch} of the original
data.} to show that SGHMC maintains the desired target distribution as the stationary distribution.
Our methods are similar in that we use mini-batches of data, but the difference is that they focus
on improving a particular flavor of MCMC, \emph{Hamiltonian} Monte Carlo (which uses Langevin
dynamics to add noise), and they also try avoiding MH tests, while we are focused on improving
\emph{general} MCMC by means of a \emph{more efficient} MH proposal. These are different ways of
solving the same problem (i.e., the problem of using mini-batches of large data for MCMC).

The work of~\cite{cutting_mh_2014} also concerns speeding up MCMC by using mini-batches of data.
Their argument is that the standard MCMC algorithm for Bayesian posterior inference (i.e., to
compute $p(\theta \mid x_1, \ldots, x_N)$ for i.i.d. data points) is unsuitable for large datasets
because the likelihood computation requires the use of all points, $p(x_1, \ldots, x_n \mid \theta)
= \prod_{i=1}^Np(x_i\mid \theta)$. Using all of this data just to decide whether the new sample
$\theta$ should be accepted or rejected is computationally inefficient. In theory, given infinite
time, using all the data points is fine because it results in an unbiased estimate of the target
distribution and the variance of the resulting sampled values is low given enough
time.\footnote{Note that~\cite{cutting_mh_2014} primarily use MCMC to estimate just the expectation
of the target distribution, but their analysis should extend to other values of the target
distribution for arbitrary input $\theta$. In other words, they use MCMC to estimate
$\mathbb{E}[p(\theta \mid x_1, \ldots, x_N)]$, but we are mostly interested in $p(\theta \mid x_1,
\ldots, x_N)$.} In reality, we have limited time. The key argument in~\cite{cutting_mh_2014} is that
given a fixed budget, it may be better to use $n$ data points per iteration (where $n \ll N$), which
will sacrifice some bias but will perhaps improve variance by allowing us to sample for $\theta$
more often, since we will get $\theta_1, \ldots, \theta_T$ as opposed to the standard case using all
$N$ samples but with only $\theta_1, \ldots, \theta_t$ samples ($t \ll T$).

More formally, the algorithm they develop is a sequential hypothesis testing algorithm. During each
iteration, their algorithm generates a new sample $\theta'$ from the proposal distribution. They
start with a small mini-batch of the data, and test the hypothesis (based on this small mini-batch)
that $\theta'$ should be accepted or rejected based the likelihood ratio. (Intuitively, we can
easily tell if $\theta'$ will be accepted/rejected if the mini-batch of data results in likelihood
drastically different from the likelihood of the current sample $\theta$.) If the hypothesis test is
too uncertain, then we increase the mini-batch size and apply the test again, and we repeat this
increment until we accept or reject $\theta'$ with high confidence (based on a Student-t
distribution). The downside of this algorithm is the need to keep incrementing the mini-batch size
in one iteration; in the worst case, they may not be able to tell with confidence to accept or
reject until we get nearly all $N$ of the data points, which defeats the purpose of the algorithm.
We propose our own algorithm that avoids this problem. {\color{blue} Daniel: does this make sense as
a ``weakness'' of their approach?  Ideally I would like to have a comparison between our method and
theirs.}

Some additional references to use:

\begin{itemize}
    \item \cite{icml2014c1_bardenet14} (ICML 2014) I think we definitely want to consider this. It
    seems to be ``an alternative to the Cutting the MH budget approach'' and so it might also be
    considered as a direct competitor.
    \item \cite{conf/icml/AhnBW12} (ICML 2012) is the ``successor'' to Stochastic Gradient Langevin
    Dynamics.
    \item \cite{conf/uai/MaclaurinA14} (UAI 2014) seems like it would be interesting to state its
    limitations.
    \item \cite{mcmc_hamiltonian_2010} is a long paper and is written like a chapter in a textbook,
    so it's for pedagogical reasons. It describes MCMC with Hamiltonian dynamics. It is useful to
    gain some background on a lot of the HMC-related papers that have come since its publication
    (2010).
    \item \cite{stochastic_thermostats_2014} (NIPS 2014) might be considered as a ``competitor''
    though only for HMC cases.
    \item \cite{sgmcmc_2015} (NIPS 2015) might want to give this a mention, where it would fit in
    their recipe.
\end{itemize}

\section{Our Algorithm}\label{sec:our_algorithm}

\subsection{The General MCMC Method}\label{sec:general_mcmc}

The general MCMC method proceeds as follows. For a desired random vector $\theta$, we have a
\emph{target distribution} we wish to compute, $p(\theta \mid x_1, \ldots, x_N)$, based on $N$
i.i.d. data points. Since exact evaluation is intractable, we generate a chain of correlated samples
$\theta_1, \ldots, \theta_T$ for some large $T$, and approximate $p$ by using counts of samples. For
each iteration $t$, we start with our current $\theta_t$. We use a \emph{proposal distribution}
$q(\theta' \mid \theta_t)$ to determine a new candidate $\theta'$. With probability $P_a$, we accept
it and set $\theta_{t+1} = \theta'$; otherwise, we repeat the previous value $\theta_{t+1} =
\theta_t$. Traditionally, $P_a$ is computed as follows:
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i ; \theta')q(\theta_t \mid \theta')}{p(\theta_t)\prod_{i=1}^N p(x_i ; \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t|x_1,\ldots,x_N) \propto p(\theta_t)\prod_{i=1}^N p(x_i
\mid \theta_t)$ which roughly tells us how ``likely'' we are to have a $\theta_t$ sample. The
normalizing constants cancel out, so the requirement for $P_a$ is that $f$ be proportional to the
true target distribution.

One then draws a uniform random variable $u$ from ${\rm Unif}[0,1]$ and accepts $\theta_{t+1} =
\theta'$ if $u < P_a$, and uses $\theta_{t+1} = \theta_t$ otherwise. The $P_a$ from
Equation~\ref{eq:traditional} satisfies ``detailed balance'' which, roughly speaking, means that if
we sample long enough using the heuristic above, we will arrive at a stationary distribution.

The $P_a$ is guaranteed to converge to the true $p$ distribution given sufficiently many examples
(though a burn-in period and/or taking every $n$th sample may be necessary). Unfortunately,
computing $f$ requires the use of all $N$ training data points. In this paper, we derive a new fast,
approximate Metropolis-Hastings test which does not use all $N$ points during each test.

\subsection{A Different Metropolis-Hastings Test}\label{sec:faster}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{john_bair_fig02}
  \caption{
  To the left, we have $\Delta U$ (which is simplified as $\Delta$ in
  Section~\ref{sec:our_algorithm}), along with two noise terms that come from the new variable we
  define $X$. The details are in Section~\ref{sec:our_algorithm}.
  }
  \label{fig:part2}
\end{figure}

@@

\section{Implementation}\label{sec:implementation}

>>>>

\title{A New Metropolis-Hastings Test for Fast Minibatch MCMC Inference and Optimization}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods are one of the most popular techniques in machine learning
for Bayesian inference. With large datasets, however, MCMC methods can be intractably slow due to
the need to compute the likelihood of all data points each iteration. Borrowing from the
optimization field, one way to make MCMC faster is to compute the posterior based on a random subset
of the data each iteration. Such stochastic MCMC has received increased research attention in recent
years, but these methods generally require adaptively increasing the minibatch size each iteration,
which can nullify the benefits if too much data is needed. In addition, with large datasets and
model optimization, it is unclear if obtaining a full posterior distribution is desirable over
getting a high-quality point MAP or ML estimate of the parameters.  In this paper, we present a new
Metropolis-Hastings test to fit in our alternative perspective on stochastic MCMC methods for big
data model optimization. We demonstrate the benefits of our approach with theoretical and empirical
results.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is one of the most popular techniques for Bayesian
computation. MCMC algorithms propose samples from a proposal distribution $q$, and decide whether to
accept or reject them based on a Metropolis-Hastings test~\cite{Metropolis1953,hastings70}. With
traditional Bayesian posterior inference, the posterior distribution $p(\theta \mid x_1, \ldots,
x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ of parameter $\theta$ based on conditionally
independent data $\{x_i\}_{i=1}^n$ is the target distribution which one aims to approximate by
sampling $\theta$ values. When $n$ is large (e.g., in the billions), it is computationally difficult
to compute the exact posterior because the likelihood must be evaluated on each $x_i$.

Recent success in machine learning optimization in the presence of large datasets have been in part
due to the ubiquitous use of stochastic gradient methods, where a subset of the data is used each
iteration to compute a noisy estimate of the gradient. These Robbins-Munro
algorithms~\cite{RobbinsMonro1951} have historically not been used with MCMC methods. A recent
result from~\cite{cutting_mh_2014} attempted to bridge the gap between minibatch optimization and
Bayesian inference by proposing an \emph{adaptive minibatch} method. Their method uses a subset of
the $N$ data points for the MH test each iteration, but requires incrementally increasing the
minibatch size so that it is more confident in its prediction on accepting or rejecting.

We argue for an alternative approach to using MCMC methods in the context of big data problems. In
general, model optimization involves finding one point estimate of optimal parameters based on a
user's goals, which typically refer to MAP or ML estimates. In addition, the more data one has, the
sharper the posterior distribution's modes. In the extreme case, the posterior would collapse to the
points corresponding to the modes, and it would be difficult for MCMC methods to escape.

Therefore, we are not interested in exact inference of the posterior model parameters for a very
large dataset, but rather, we want to perform either MAP or ML estimation. In both cases the goal is
to find a good posterior mode, but the exact form of the distribution is arbitrary. In particular,
non-unit powers of the distribution are fine. That corresponds to changing the temperature of the
distribution to make it ``flatter''.

Rather than fixing the posterior temperature and trying to use adjustable mini-batch size to achieve
it (which ends up reducing to nearly-full-batch updates) we propose to run at the natural
temperature induced by a given minibatch size. We can always increase the minibatch size to reduce
the temperature, but there is no need to do this beyond the point where the posterior mode is sharp
and well-defined. That will occur at some fixed number of points and far less than the full dataset.

To this end, we present a new Metropolis-Hastings test that can fit in this framework of minibatch
MCMC methods. We present theoretical and experimental results on our method and show that it can be
the method of choice for minibatch MCMC .

\section{Preliminaries and Related Work}\label{sec:related_work}

The standard MCMC method proceeds as follows~\cite{gilks1996markov,brooks2011handbook}. For a
parameter $\theta$ and conditionally independent data, we wish to compute the target distribution
$p(\theta \mid x_1, \ldots, x_N)$. Since exact evaluation is intractable, we generate a chain of
correlated samples $\theta_1, \ldots, \theta_T$ for some large $T$, and approximate $p$ by using
counts of samples. For each iteration $t$, we start with our current $\theta_t$. We use a
\emph{proposal distribution} $q(\theta' \mid \theta_t)$ to determine a new candidate $\theta'$. With
probability $P_a$, we accept it and set $\theta_{t+1} = \theta'$; otherwise, we repeat the previous
value $\theta_{t+1} = \theta_t$. Traditionally, $P_a$ is computed as follows:
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i ; \theta')q(\theta_t \mid \theta')}{p(\theta_t)\prod_{i=1}^N p(x_i ; \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t|x_1,\ldots,x_N) \propto p(\theta_t)\prod_{i=1}^N p(x_i \mid
\theta_t)$. The technical requirement for $P_a$ is that $f$ be proportional to the true target
distribution.

One then draws a uniform random variable $u$ from ${\rm Unif}[0,1]$ and accepts $\theta_{t+1} =
\theta'$ if $u < P_a$, and uses $\theta_{t+1} = \theta_t$ otherwise. The $P_a$ from
Equation~\ref{eq:traditional} satisfies detailed balance, so if one samples long enough using the
heuristic above, one will arrive at a stationary distribution.  The $P_a$ is guaranteed to converge
to the true $p$ distribution given sufficiently many examples (though a burn-in period and/or taking
every $n$th sample may be necessary).

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. There has been research on ways to conduct exact accept/reject tests using unbiased
estimators of the likelihood~\cite{Andrieu09thepseudo-marginal}. However, unbiased estimators of the
likelihood that can be computed from minibatches of data, such as the Kennedy-Bhanot
estimator~\cite{PhysRevD} have very high variance for large datasets. Because of this, once we get a
very high estimate of the likelihood, almost all proposed moves are rejected and the algorithm gets
stuck.

{\color{blue}
Daniel: I am worried. We need to motivate why it is not easy to use minibatch MCMC, and what other
researchers have tried to do about that.  The last paragraph is from~\cite{cutting_mh_2014}, and it
seems to argue that people have not used minibatch MCMC before ``the algorithm gets stuck''. But
isn't that exactly what our algorithm is doing when it is getting ``stuck'' at modes? If we wanted
to get stuck at modes with large datasets, why didn't we just do what previous work did?
}

The work of~\cite{cutting_mh_2014} proposes an adaptive minibatch approach by using a sequential
hypothesis testing algorithm. During each iteration, they start with a small minibatch of data and
test the hypothesis that the sample $\theta'$ should be accepted or rejected based the likelihood
ratio. If the test cannot make a decision over a certain confidence threshold, then they increase
the minibatch size and test again. This process repeats until a decision.  The downside of this
algorithm is the need to keep incrementing the mini-batch size in one iteration; in the worst case,
all $N$ data points may be needed in a single iteration. A related approach
is~\cite{icml2014c1_bardenet14}, which has slightly more robust theoretical properties, but comes at
the cost of more computational time. The work of~\cite{conf/uai/MaclaurinA14} presents an
\emph{exact} approach, but has the sometimes unreasonable requirement of a lower bound on the
per-datum likelihood factors.

There is a separate line of work related to optimizing MCMC about simulating the physics of a
probability distribution. By viewing random variables as ``particles'' in a system, it is possible
to apply Hamiltonian Monte Carlo (HMC)~\cite{mcmc_hamiltonian_2010} methods which generate both high
quality and distant proposals, getting the best of both worlds. These require a full gradient
computation of the posterior, which means they also suffer from the problem of needing to know the
likelihood at each point. Stochastic versions of HMC~\cite{sghmc_2014,stochastic_thermostats_2014}
use a minibatch of data each iteration but require controlling the noise from sampling so the sample
sequence does not diverge. A similar approach is to use Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12}, which does not use the momentum terms in HMC.

Ultimately, minibatch SGHMC and Langevin Dynamics is an orthogonal contribution to ours, two ways of
attempting to solve the same problem of MCMC methods with big datasets. We could augment their
methods with our MH test, to get the benefit of their high quality proposals. Note that while their
methods in theory do not need an MH test, in practice, they need one due to discretization of the
system. Furthermore, tests may also be needed for when the algorithms are run with a constant step
size.

%% Daniel: I don't know how to integrate this easily.
%% \cite{sgmcmc_2015} (NIPS 2015) might want to give this a mention, where it would fit in their
%% recipe.

\section{A New Metropolis-Hastings Test}\label{sec:our_algorithm}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{john_bair_fig02}
  \caption{
  To the left, we have $\Delta U$ (which is simplified as $\Delta$ in
  Section~\ref{sec:our_algorithm}), along with two noise terms that come from the new variable we
  define $X$. The details are in Section~\ref{sec:our_algorithm}.
  {\color{blue}
  Daniel: I think this needs to be merged into Figure~\ref{fig:part1}.
  }
  }
  \label{fig:part2}
\end{figure}

@@

\section{Theoretical Results}\label{sec:theory}

TODO insert Haoyu's proof.
